Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)\thanks{Code and data are available at: \url{https://github.com/wendyhuan/lifeexpectancy}.}

### 2024-12-02 paper/paper.tex
latex

@@

To examine the differences in life expectancy under the same variables,
we separate the data into developed and developing countries, where
developed country are generally believed to have higher life expectancy.
Figure~\ref{fig-expectancy} displays density of the historical data of
life expectancy across 195 countries from 2000 to 2015, comparing
developing to developed countries.The average age of most developed
country is approximately 80, while the majority developing countries
typically have average life expectancy around mid-seventies.

@@

\caption{\label{fig-expectancy}This histogram shows the distribution of
the percentage of life expectancy.}

@@

We began by importing the raw dataset using the read\_csv function from
the tidyverse package. To focus our analysis on more relevant variables,
we selected specific columns, such as percentage of expenditure of the
country, total health expenditure by government .etc., omitting any
unnecessary columns.

We filter out the rows containing NA values in any of the selected
columns for reducing the noise and simpler further analysis..

We then renaming columns for clarity. For example, we changed
`Income.composition.of.resources' into `IncomeComposition' , making it
easier for anyone working with the data to read and understand what each
variable represents.

Each column is rounded using the round function, specifying the desired
number of decimal places for each. Columns not mentioned in mutate
remain unchanged. This ensures a flexible and precise cleaning process
tailored to further compairation and graphing.

We also modified the country name of `Republic of Korea' under Country
variable to `South Korea' to avoid misunderstanding of North Korea.

We also created a new variable called Continent, which indicates which
continent does the country comes from in order to provide a geographical
context to the analysis. Life expectancy may be higher in developed
regions like Europe or North America compared to regions like
Sub-Saharan Africa due to differences in healthcare, living standards,
and economic development.

We also wrapped the cleaning process in a tryCatch block in order to
mitigate any errors that arose throughout the cleaning process.

>>>>

\textbf{?@fig-cleaned} shows the average of life expectancy under male,
female, 6 different regions and 4 income groups.

@@

Figure~\ref{fig-expectancy} visualizes the distribution of life
expectancy from 2015 to 2020 as percentages, combining a histogram and a
density plot. A large concentration of observations is observed range of
0\% to 25\%, reflecting a significant proportion of countries with lower
life expectancy relative to others in the dataset. Another visible
concentration is around 75\% to 100\%, indicating a group of countries
with relatively high life expectancy. The blue density curve overlays
the histogram, providing a smoothed representation of the distribution.
Peaks in the curve correspond to high frequency within the histogram.
The density plot emphasized the bimodal nature of the distribution with
two distinct peaks: one around 12 as low life expectancy percentages;
another peak around 87as high life expectancy percentage.

@@

\caption{\label{fig-expectancy}This histogram shows the distribution of
the percentage of life expectancy. With a huge gap between 25 and 50, we
could indicate that most of the countries has a life expectancy of.
There exist large data with low life expectancy less than 25 years.}

@@

We began by importing the raw dataset using the read\_csv function from
the tidyverse package. To focus our analysis on more relevant variables,
we selected specific columns, such as Indicator of life expectancy as
birth and life expectancy at 60 years, Country, Period, Gender, and Life
expectancy, omitting any unnecessary columns.

We functioned to extract the number before opening the square bracket of
life expectancy since the raw data gave both the range and the mean of
life expectancy of each country under each year.

We filter out the rows containing NA values in any of the selected
columns for reducing the noise and simpler further analysis.

We then renaming columns for clarity. For example, we changed `Location'
into `Country', ``Dim1'' for ``Gender'' and ``Value'' for ``Life
Expectancy'', making it easier for anyone working with the data to read
and understand what each variable represents.

Each column is rounded using the round function, specifying the desired
number of decimal places for each. Columns not mentioned in mutate
remain unchanged. This ensures a flexible and precise cleaning process
tailored to further comparison and graphing. We dropped the potential
percentage symbol to contain purely numerical data.

The income group column and the region column was not given in the raw
dataset. We created a new variable called ``Income\_Group'' by mapping
over country name to four income group (lower\_income, lower\_middle,
upper\_middle, high\_income) according to the index given by World Bank
Income Group. This is a key variable in predicting life expectancy since
it is believed that life expectancy is correlated to the income
situation. We as well made a mapping to the countries according to its
continent and saved as ``Region''. This is the variable indicates the
geographic context to the analysis. Countries are mapped into six
continents (``Africa'', ``Asia'', ``Europe'', ``North America'', ``South
America'', ``Oceania'') one by one.

We again merge the table, removing repeated columns and rows and mapping
the region and the country income under different given names.

Lastly, we filter out NAs again as a further check. We continue filter
out the Palestinian territory, which is not a country and was not mapped
in neither Region nor Income Group.

For easy reading format, we arrange the country into Alphabet order and
mutate the variables again for converting columns into appropriate data
types.

For easier visualization, we created four tables, which are grouped by
life expectancy at age 60, life expectancy at birth, life expectancy at
birth of Male and life expectancy at birth of Female.
